A comparison on R tools for nonlinear least squares modeling

### 2023-05-02 RefactoringNLS/ComparingNLStools.Rmd
trying to streamline

The tools we will consider are the base-R `nls()` function 
and a number of packages that are available on the CRAN repository.
Of these, we will pay particular attention to **nlsr** (@nlsr2023manual),
**minpack.lm** (@minpacklm12), gslnls (@gslnls23), and nls2 (@nls2-22).

@@

The tools for nonlinear estimation we consider have been built with different
goals and contexts. We will make our comparison in a number of directions.
We will make a number of specific notes, but our main commentary will consider
the following categories. These overlap, for example in that the emphasis on modeling
versus solving the nonlinear least squares problem will result in different
output.f

@@

The goal of many statisticians and other workers is to estimate a model. The
minimization of a sum of squared residuals -- the nonlinear least squares problem --
is the mechanism to estimate the model. Some of the tools we review focus on the
modeling, so attempt to provide measures of uncertainty and other information. 
The interface to modeling tools will generally try to streamline that task. By
contrast, nonlinear least squares tools are trying to solve the more limited
numerical task.

@@

The necessarily iterative solution of a nonlinear least squares problem 
to carry out nonlinear modeling requires that we can find smaller values
of a sum of squared residuals. This implies there is a "downhill" direction.
The gradient of the residual sum of squares is computed as the product
of the Jacobian matrix: the partial derivatives of the residual vector 
with respect to the modeling parameters.

Most researchers find the effort to interface the differential calculus 
required to specify the Jacobian to a nonlinear estimation tool very 
challenging. The work-around is generally to use a numerical approximation,
but there is the possibility of using symbolic or automatic differentiation,
and there are particular models where workers have provided code for the
Jacobian.

@@

Reality often imposes conditions on parameters. While general constraints can
be difficult to include, bounds on parameters can be accommodated by many 
solvers. For some reason, fixed parameters (masks) are not often considered,
but are very convenient when a parameter has a commonly accepted value that we
later wish to allow to be estimated from the data.

@@

## Other CRAN packages

`onls` (@R-onls-manual) -- The usual optimization in estimating nonlinear models
is the vertical difference between the dependent variable and the functional
model. `onls` minimizes the sum of orthogonal residuals. The objective is
therefore different, and involves nontrivial extra calculation. A vignette with
the package and the blog article 
https://www.r-bloggers.com/2015/01/introducing-orthogonal-nonlinear-least-squares-regression-in-r/ 
give some description with illustrative graphs. `onls` appears to be limited to 
problems with one independent and one dependent variable. The Wikipedia article
https://en.wikipedia.org/wiki/Total_least_squares presents an overview of some
ideas, with references to the literature. Our view is that the approach needs wider 
discussion and tutorial examples to allow its merits to be judged.

>>>>

The tools we will consider are the base-R `nls()` function 
and a number of packages that are available on the CRAN repository.
Of these, we will pay particular attention to **nlsr** (@nlsr2023manual),
**minpack.lm** (@minpacklm12), **gslnls** (@gslnls23), and **nls2** (@nls2-22).

@@

The tools for nonlinear estimation we consider have been built with different
goals and contexts. We will make our comparison considering the overlapping
headings summarized below.

@@

The goal of many statisticians and other workers is to estimate a model. The
minimization of a sum of squared residuals -- the nonlinear least squares problem --
is the mechanism to estimate the model. Some tools focus on modeling, 
so attempt to provide measures of uncertainty and other information. 
The interface to modeling tools will generally try to streamline that task. By
contrast, nonlinear least squares tools are trying to solve the more limited
numerical task.

@@

The necessarily iterative solution of a nonlinear least squares problem 
to carry out nonlinear modeling requires that we can find smaller values
of a sum of squared residuals. This implies there is a "downhill" direction.
The gradient of the residual sum of squares is computed as its product
with the Jacobian matrix: the partial derivatives of the residual vector 
with respect to the modeling parameters.

Most researchers are challenged to program the differential calculus 
required to specify the Jacobian to a nonlinear estimation tool. 
The work-around is generally to use a numerical approximation,
but there is the possibility of using symbolic or automatic differentiation.
There are also particular models where workers have provided code for the
Jacobian.

@@

Reality imposes conditions on parameters. While general constraints can
be difficult to include, bounds on parameters can be accommodated by many 
solvers. For some reason, fixed parameters (masks) are not often considered,
but are convenient when a parameter has a commonly accepted value that we
later wish to allow to be estimated from the data.

@@

## Other CRAN packages considered

`onls` (@R-onls-manual) -- The usual optimization in estimating nonlinear models
is the vertical difference between the dependent variable and the functional
model. `onls` minimizes the sum of **orthogonal** residuals. The objective is
therefore different, and involves nontrivial extra calculation. A vignette with
the package and the blog article 
https://www.r-bloggers.com/2015/01/introducing-orthogonal-nonlinear-least-squares-regression-in-r/ 
give some description with illustrative graphs. `onls` appears to be limited to 
problems with one independent and one dependent variable. The Wikipedia article
https://en.wikipedia.org/wiki/Total_least_squares presents an overview of some
ideas, with references to the literature. Our view is that the approach needs wider 
discussion and tutorial examples to allow its merits to be judged.
